The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-08 rational.tex
intro

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! eta = \eta
%! eps = \epsilon
%! eta+ = \eta^+
%! eps+ = \epsilon^+
%! eta* = \eta^{\times}
%! eps* = \epsilon^{\times}
%! trace+ = trace^+
%! trace* = trace^{\times}
%! ^^^ = ^{-1}
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +
%! LeftP = L^+
%! RightP = R^+
%! LeftT = L^{\times}
%! RightT = R^{\times}

@@

To get a feeling for the main idea, consider the following algebraic
manipulation relating a natural number $a$ to itself:
\[\begin{array}{rclll}
&& a & \mbox{initial~$a$} & (0) \\
&=& a + 0 & \mbox{$0$~is~identity~for~$+$} & (1) \\
&=& a + (-a + a) & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (2) \\
&=& (a + (-a)) + a & \mbox{$+$~is~associative} & (3) \\
&=& 0 + a & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (4) \\
&=& a & \mbox{$0$~is~identity~for~$+$} & (5)
\end{array}\]
Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism of type {{a <-> a}} with a non-trivial and interesting
computational interpretation. Consider the starting value of type $a$ as
being supplied by a producer of values and the final value of type $a$ as
being demanded by a consumer. For the purposes of the example, think of the
producer as someone having \$20.00 and the consumer as someone demanding
\$20.00. It is certainly possible to directly connect the producer and
consumer using the algebraic identity $a=a$: such a connection would be
computationally modeled by the identity function as expected. However, the
algebraic proof above implements this connection in a much more interesting
way. As the semantics of Sec.~\ref{sec:rat} formalizes, the flow of 
computation is as follows:
\begin{itemize}
\item We start at line (0) with \$20.00; 
\item We proceed to line (1) with the same \$20.00 but tagged as being in the
  left summand of the type $a+0$; we indicate this value as {{left 20}}
\item We continue to line (2) with the same value {{left 20}};
\item At line (3), the tag on the \$20.00 changes to indicate that it is
  in the left-left summand, i.e., the value is now {{left (left 20)}};
\item At line (4), we find ourselves needing to produce a value of type 0
  which is impossible; this signals the beginning of a reverse execution
  which sends us back to line (3) with a value {{left (right 20)}};
\item Execution continues in reverse to line (2) with the value 
  {{right (left 20)}};
\item At line (1) we find ourselves again facing an empty type so we reverse
  execution again; we go to line (2) with a value {{right (right 20)}};
\item We proceed to line (3) and (4) with the value {{right 20}};
\item We finally reach line 5 with the value {{20}}.
\end{itemize}
The beauty of the above computation is that it can be realized with a
completely different concurrent schedule as follows. \emph{Simultaneously}
the execution begins at line (0) with the value {{20}}, at line (2) with the
value {{right (right 20)}} flowing forward and the value {{right (left 20)}}
flowing backwards, and at line (3) with the value {{left (left 20)}} flowing
forward and the value {{left (right 20)}} flowing backward. We now have three
independent subcomputations proceeding in parallel: the positive value
created at line (2) which can flow to the consumer; the negative value
created at line (3) which can pay for the debt created at line (2); and the
initial value from line (0) which can flow to the required forward-flowing
value needed at line (3). At some intuitive level, one might understand the
situation as modeling a credit card payment in which money is created at line
(2) for the consumer together with a corresponding debt. The producer's bank
pays for the debt at line (3) and waits for the actual money to arrive from
the producer.

>>>>

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! eta = \eta
%! eps = \epsilon
%! eta+ = \eta^+
%! eps+ = \epsilon^+
%! eta* = \eta^{\times}
%! eps* = \epsilon^{\times}
%! trace+ = trace^+
%! trace* = trace^{\times}
%! ^^^ = ^{-1}
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +
%! LeftP = L^+
%! RightP = R^+
%! LeftT = L^{\times}
%! RightT = R^{\times}
%! alpha = \alpha

@@

To get a feeling for the main idea, consider the following algebraic
manipulation relating a natural number $a$ to itself:
\[\begin{array}{rclll}
&& a & \mbox{initial~$a$} & (0) \\
&=& a + 0 & \mbox{$0$~is~identity~for~$+$} & (1) \\
&=& a + (-a + a) & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (2) \\
&=& (a + (-a)) + a & \mbox{$+$~is~associative} & (3) \\
&=& 0 + a & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (4) \\
&=& a & \mbox{$0$~is~identity~for~$+$} & (5)
\end{array}\]
Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism of type {{a <-> a}} with a non-trivial and interesting
computational interpretation. The witness for this isomorphism is a
computation that takes a value of type $a$ (say \$20.00) and eventually
produces the same \$20.00 value as its output. As the semantics of
Sec.~\ref{sec:rat} formalizes, the intermediate steps of the computation
proceed as follows:
\begin{itemize}
\item We start at line (0) with \$20.00; 
\item We proceed to line (1) with the same \$20.00 but tagged as being in the
  left summand of the type $a+0$; we indicate this value as {{left 20}}
\item We continue to line (2) with the same value {{left 20}};
\item At line (3), as a result of re-association the tag on the \$20.00
  changes to indicate that it is in the left-left summand, i.e., the value is
  now {{left (left 20)}};
\item At line (4), we find ourselves needing to produce a value of type 0
  which is impossible; this signals the beginning of a reverse execution
  which sends us back to line (3) with a value {{left (right 20)}};
\item Execution continues in reverse to line (2) with the value 
  {{right (left 20)}};
\item At line (1) we find ourselves again facing an empty type so we reverse
  execution again; we go to line (2) with a value {{right (right 20)}};
\item We proceed to line (3) and (4) with the value {{right 20}};
\item We finally reach line 5 with the value {{20}}.
\end{itemize}

Now consider the following algebraic manipulation (assuming $a \neq 0$):
\[\begin{array}{rclll}
&& a & \mbox{initial~$a$} & (0) \\
&=& a * 1 & \mbox{$1$~is~identity~for~$*$} & (1) \\
&=& a * ((1/a) * a) & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (2) \\
&=& (a * (1/a)) * a & \mbox{$*$~is~associative} & (3) \\
&=& 1 * a & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (4) \\
&=& a & \mbox{$1$~is~identity~for~$*$} & (5)
\end{array}\]
As the semantics in Sec.~\ref{sec:rat} formalizes, the witness to this 
{{a <-> a}} isomorphism commputes as follows:
\begin{itemize}
\item We start at line (0) with the value {{20}};
\item We proceed to line (1) with the value {{(20,())}};
\item We proceed to line (2) with the value {{(20,((1/alpha),alpha))}} where
  $\alpha$ is essentially a logic variable;
\item We proceed to line (3) with the value {{((20,1/alpha),alpha)}};
\item At line (4) we require that $20$ and $1/\alpha$ cancel each other which
  forces the unification $\alpha=20$, i.e., the value becomes
  {{((20,1/20),20)}} which simplifies to {{((),20)}};
\item We reach line (5) with the value {{20}}.
\end{itemize}

\amr{the above is a boring trace; we need to convey why it is exciting but we
  are not agreeing on the interpretation.}

The beauty of the above computation is that it can be realized with a
completely different concurrent schedule as follows. \emph{Simultaneously}
the execution begins at line (0) with the value {{20}}, at line (2) with the
value {{right (right 20)}} flowing forward and the value {{right (left 20)}}
flowing backwards, and at line (3) with the value {{left (left 20)}} flowing
forward and the value {{left (right 20)}} flowing backward. We now have three
independent subcomputations proceeding in parallel: the positive value
created at line (2) which can flow to the consumer; the negative value
created at line (3) which can pay for the debt created at line (2); and the
initial value from line (0) which can flow to the required forward-flowing
value needed at line (3). At some intuitive level, one might understand the
situation as modeling a credit card payment in which money is created at line
(2) for the consumer together with a corresponding debt. The producer's bank
pays for the debt at line (3) and waits for the actual money to arrive from
the producer.
